---
title: "Stat501_Homework6"
author: "Kelby Kies"
date: "4/7/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1: Consider the crabs dataset in R used in Exam1.

## part A.) Use principal components analysis to reduce the dimensionality of the crabs dataset into two dimensions.Display the results. Is there any distinctiveness in the four species/sex combinations? [10 points]
```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(MASS)
crabs <- crabs
crabs$species <- as.factor(paste0(crabs$sp,crabs$sex))


pca1 <- prcomp(crabs[,-c(1,2,3,9)], scale = T)
print("PCA SUMMARY:")
summary(pca1)
plot(pca1$x[,1:2], pch=16, col=as.numeric(as.factor(crabs[, "species"])), main="Crab Data PCA separated by Species and Sex") 
legend(3, 0.75, pch=16, col=unique(as.numeric(as.factor(crabs[, "species"]))),legend=unique(crabs[, "species"]))


# Remove Group Sepcific Means
# crabs.centered <- log(crabs[,4:8]) - rbind(matrix(apply(log(crabs[crabs[,9]=="BM",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="BM"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="BF",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="BF"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="OM",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="OM"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="OF",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="OF"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T))
# Calculate PCA after removing the group specific Means
# We are removing the means here because we know that the different 
#groups have different means, but the same var-cav
#pca2 <- prcomp(crabs.centered)

#summary(pca2)
#plot(pca2$x[,1:2], pch=16, col=as.numeric(as.factor(crabs[, "species"])), main="Crab Data PCA separated by Species and Sex") 
# t <- pca2$x + rbind(matrix(apply(log(crabs[crabs[,9]=="BM",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="BM"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="BF",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="BF"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="OM",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="OM"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="OF",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="OF"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T)) %*% pca$rotation
# plot(t[,1:2], pch=16, col=as.numeric(as.factor(crabs[, "species"])), main="Crab Data PCA separated by Species and Sex") 
# legend(1, 4, pch=16, col=unique(as.numeric(as.factor(crabs[, "species"]))),legend=unique(crabs[, "species"]))

# pairs(pca$x + rbind(matrix(apply(log(crabs[crabs[,9]=="BM",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="BM"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="BF",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="BF"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="OM",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="OM"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T),
#                                  matrix(apply(log(crabs[crabs[,9]=="OF",4:8]),2,mean),
#                                                 nrow = sum(crabs[,9]=="OF"),
#                                                 ncol = ncol(crabs[,4:8]),
#                                                 by = T)) %*% pca$rotation,
#       panel=function(x,y){panel.smooth(x,y, col = as.numeric(crabs[,9])+2,
#         pch = 20, cex = 2)
#                           abline(lsfit(x,y),lty=2) })

```
Is there any distinctiveness in the four species/sex combinations? [10 points]
PC1 accounts for ~98% of the variance in the data. When we account for PC2 we are looking at 99% of the total variation in the data. The biggest distinction is between Blue/Males, Orange/Males and Orange/Females crabs, thus shows a difference between sex of crab rather than the color of the crab. We can see that these 4 groups are slightly clustered together, but eventually form their own clusters correlated with sex(M/F). 


## part b.) Perform a kernel principal components analysis with two features and display the results. You may use the Gaussian radial basis funcion. But display the results for different values of $\sigma = 0.2, 0.4, 0.8, 1.0, 1.5, 3$. [20 points]

```{r, echo = FALSE, warning=FALSE, message=FALSE}

library(kernlab)
sigmas <- c(0.2, 0.4, 0.8, 1, 1.5, 3) 
par(mfrow=c(2,3))

crabs[,4:8]
library(ggplot2)
for(sig in sigmas) 
  {
  kpc <- kpca(~., data=crabs[ ,4:8], kernel="rbfdot", kpar = list(sigma=sig), features=2) 
  crabs.kpc.df <- data.frame(rotated(kpc), group = paste0(crabs[,1], crabs[,2])) 
  names(crabs.kpc.df) <- c("KPC1", "KPC2", "group")
  p <- ggplot(crabs.kpc.df, aes(KPC1, KPC2, colour = group)) +
  geom_point() + theme_light()
  print(p) 
  }
kpc_2 <- kpca(~.,data=crabs[,c(-1,-2,-3,-9)],kernel="rbfdot",
                 kpar=list(sigma=0.2),features=2)
kpc_4 <- kpca(~.,data=crabs[,c(-1,-2,-9)],kernel="rbfdot",
                 kpar=list(sigma=0.4),features=2)
kpc_8 <- kpca(~.,data=crabs[,c(-1,-2,-9)],kernel="rbfdot",
                 kpar=list(sigma=0.8),features=2)
kpc_1 <- kpca(~.,data=crabs[,c(-1,-2,-9)],kernel="rbfdot",
                 kpar=list(sigma=1.0),features=2)
kpc_15 <- kpca(~.,data=crabs[,c(-1,-2,-9)],kernel="rbfdot",
                 kpar=list(sigma=1.5),features=2)
kpc_3 <- kpca(~.,data=crabs[,c(-1,-2,-9)],kernel="rbfdot",
                 kpar=list(sigma=3),features=2)

#plot the data projection on the components

plot(rotated(kpc_2),col=as.numeric(as.factor(crabs[,9])), xlab="1st Principal Component",ylab="2nd Principal Component", main ="Crab Data Kernel PCA separated by Species and Sex: sigma = 0.2")
legend("topright", pch=16, col=unique(as.numeric(as.factor(crabs[,9]))),legend=unique(crabs[,9]))

plot(rotated(kpc_4),col=as.numeric(as.factor(crabs[,9])), xlab="1st Principal Component",ylab="2nd Principal Component", main ="Crab Data Kernel PCA separated by Species and Sex: sigma = 0.4")
legend("topleft", pch=16, col=unique(as.numeric(as.factor(crabs[,9]))),legend=unique(crabs[,9]))

plot(rotated(kpc_8),col=as.numeric(as.factor(crabs[,9])), xlab="1st Principal Component",ylab="2nd Principal Component", main ="Crab Data Kernel PCA separated by Species and Sex: sigma = 0.8")
legend("topright", pch=16, col=unique(as.numeric(as.factor(crabs[,9]))),legend=unique(crabs[,9]))

plot(rotated(kpc_1),col=as.numeric(as.factor(crabs[,9])), xlab="1st Principal Component",ylab="2nd Principal Component", main ="Crab Data Kernel PCA separated by Species and Sex: sigma = 1")
legend("topleft", pch=16, col=unique(as.numeric(as.factor(crabs[,9]))),legend=unique(crabs[,9]))

plot(rotated(kpc_15),col=as.numeric(as.factor(crabs[,9])), xlab="1st Principal Component",ylab="2nd Principal Component", main ="Crab Data Kernel PCA separated by Species and Sex: sigma = 1.5")
legend("bottomleft", pch=16, col=unique(as.numeric(as.factor(crabs[,9]))),legend=unique(crabs[,9]))

plot(rotated(kpc_3),col=as.numeric(as.factor(crabs[,9])), xlab="1st Principal Component",ylab="2nd Principal Component", main ="Crab Data Kernel PCA separated by Species and Sex: sigma = 3")
legend("topright", pch=16, col=unique(as.numeric(as.factor(crabs[,9]))),legend=unique(crabs[,9]))
```

# Question 2: 

## Part A.) We will see whether the digits are distinguishable. To do so, we will first prepare the dataset by rooting out those pixels (coordinates) which do not contribute to categorization. Do so, using appropriate univariate but simultaneous methods, at the 5% level. For the remainder of this question, we will be focused on the 100 most significant coordinates (in terms of the p-value for the above test). [10 points]
```{r,echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(car)
zip <- read.table('~/Desktop/stat_501/ziptrain.dat')
zip_digits <- read.table('~/Desktop/stat_501/zipdigit.dat') %>% dplyr::rename(digits = V1)

zip2 <- as.data.frame(cbind(zip_digits, zip))
#models <- list()
variables <- list()
pvals <- list()
start_df <- data.frame()
for (y in names(zip2)[-1]){
  
  form <- formula(paste(y, "~", "digits"))
  lm_model <- lm(form, data=zip2) 
  lm_anova <- anova(lm_model)
  if(lm_anova$`Pr(>F)`[1]<= 0.05){
    #models[[y]] <- lm_model
    new_df <- data.frame(y,lm_anova$`Pr(>F)`[1])
    names(new_df) <- c("var", "Pval")
    start_df <- rbind(start_df, new_df)
  }
}
end_df <- start_df[order(start_df$Pval),]

# Select the top 100 most significant variables!
final_variables <- end_df[1:100,]$var

new_zip <- dplyr::select(zip2, digits, final_variables)
print("Here are the final pixel(coordinate) variables I will use
      (Top 100 based on significane level 0.05):")
print(final_variables)
```


### part Ai.) Evaluate whether the variance-covariance matrices are all equal across all digits.[10 points]
```{r}
source('~/Desktop/stat_501/BoxMTest.R')

BoxMTest(new_zip[,-1], cl = as.factor(new_zip$digits), alpha=0.05)
```
Using the BoxMTest() function we see that the covariance matrices are significantly different across the digits when comparing at the 5% significance level. 

### part Aii.) If these are not equal, we will assume that they are the true values of the individual group dispersion matrices. Derive a likelihood ratio test statistic for testing differences in mean effects across different digits. [10 points]

See attached pdf.

## part B.) We will now use principal components to reduce dimensionality of the original dataset. Note that the images for the different digits have different means and chanracteristics, therefore, it would be preferred to remove the effect of the digit-specific means before performing the principal components analysis. (Transformed versions of these means need to be added back before proceeding much further.) Use the principal components and determine the number of components needed to explain at least 80% of the total variation in the data, at the 5% level of significance. [20 points]
```{r, echo = FALSE, warning=FALSE, message=FALSE}
zip.centered <- zip2[,-1] - rbind(matrix(apply(zip2[zip2[,1]==0,-1],2,mean),
                                                nrow = sum(zip2[,1]==0),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==1,-1],2,mean),
                                                nrow = sum(zip2[,1]==1),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==2,-1],2,mean),
                                                nrow = sum(zip2[,-1]==2),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==3,-1],2,mean),
                                                nrow = sum(zip2[,1]==3),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==4,-1],2,mean),
                                                nrow = sum(zip2[,1]==4),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==5,-1],2,mean),
                                                nrow = sum(zip2[,1]==5),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==6,-1],2,mean),
                                                nrow = sum(zip2[,-1]==6),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==7,-1],2,mean),
                                                nrow = sum(zip2[,1]==7),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==8,-1],2,mean),
                                                nrow = sum(zip2[,-1]==8),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==9,-1],2,mean),
                                                nrow = sum(zip2[,-1]==9),
                                                ncol = ncol(zip2[,-1]),
                                                by = T))

pca_zip <- prcomp(zip.centered)
print("Full PCA Summary:")
print(summary(pca_zip))
source("~/Desktop/stat_501/PCs.proportion.variation.enuff.R")
print("PC Variance Proportion at the 5% significance level")
for (i in 1:256) 
    print(c(i, PCs.proportion.variation.enuff(lambda = pca_zip$sdev^2, q = i, propn = 0.80, nobs = nrow(zip2))))

```
I first summarized the PCA I ran on the digit mean centered data. Then I used PCs.proportion.variation.enuff() to test at the 80% significance level. Both tests confirm that 80% of the total variation of the data is found within the first 32 principal components at the 5% signifcance level.  

### part B i.) Display the components (using color or characters for each digit) using appropriate methods. Compare with the displays obtained using the reduced dataset. [10 points]
```{r, echo = FALSE, warning=FALSE, message=FALSE}

y <- rbind(matrix(apply(zip2[zip2[,1]==0,-1],2,mean),
                                                nrow = sum(zip2[,1]==0),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==1,-1],2,mean),
                                                nrow = sum(zip2[,1]==1),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==2,-1],2,mean),
                                                nrow = sum(zip2[,1]==2),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==3,-1],2,mean),
                                                nrow = sum(zip2[,1]==3),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==4,-1],2,mean),
                                                nrow = sum(zip2[,1]==4),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==5,-1],2,mean),
                                                nrow = sum(zip2[,1]==5),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==6,-1],2,mean),
                                                nrow = sum(zip2[,1]==6),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==7,-1],2,mean),
                                                nrow = sum(zip2[,1]==7),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==8,-1],2,mean),
                                                nrow = sum(zip2[,1]==8),
                                                ncol = ncol(zip2[,-1]),
                                                by = T),
                                 matrix(apply(zip2[zip2[,1]==9,-1],2,mean),
                                                nrow = sum(zip2[,1]==9),
                                                ncol = ncol(zip2[,-1]),
                                                by = T)) %*% pca_zip$rotation

pairs(pca_zip$x[,1:3] + y[,1:3],
      panel=function(x,y){panel.smooth(x,y, col = as.numeric(zip2[,1])+2,
        pch = 20, cex = 1) 
        abline(lsfit(x,y),lty=2) })

# plot(pca_zip$x[,1:2], pch=16, col=as.numeric(as.factor(zip2[,"digits"])), main="Full Handwritten Zipcode PCA separated by Digit") 
# legend(12, 9, pch=16, col=unique(as.numeric(as.factor(zip2[,"digits"]))),legend=unique(zip2[,"digits"]))
```

```{r, echo = FALSE, warning=FALSE, message=FALSE}
reduced_zip_centered <- new_zip[,-1] - rbind(matrix(apply(new_zip[new_zip[,1]==0,-1],2,mean),
                                                nrow = sum(new_zip[,1]==0),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==1,-1],2,mean),
                                                nrow = sum(new_zip[,1]==1),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==2,-1],2,mean),
                                                nrow = sum(new_zip[,1]==2),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==3,-1],2,mean),
                                                nrow = sum(new_zip[,1]==3),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==4,-1],2,mean),
                                                nrow = sum(new_zip[,1]==4),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==5,-1],2,mean),
                                                nrow = sum(new_zip[,1]==5),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==6,-1],2,mean),
                                                nrow = sum(new_zip[,1]==6),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==7,-1],2,mean),
                                                nrow = sum(new_zip[,1]==7),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==8,-1],2,mean),
                                                nrow = sum(new_zip[,1]==8),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==9,-1],2,mean),
                                                nrow = sum(new_zip[,1]==9),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T)) 
reduced_zip <- prcomp(reduced_zip_centered)
z <- rbind(matrix(apply(new_zip[new_zip[,1]==0,-1],2,mean),
                                                nrow = sum(new_zip[,1]==0),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==1,-1],2,mean),
                                                nrow = sum(new_zip[,1]==1),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==2,-1],2,mean),
                                                nrow = sum(new_zip[,1]==2),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==3,-1],2,mean),
                                                nrow = sum(new_zip[,1]==3),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==4,-1],2,mean),
                                                nrow = sum(new_zip[,1]==4),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==5,-1],2,mean),
                                                nrow = sum(new_zip[,1]==5),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==6,-1],2,mean),
                                                nrow = sum(new_zip[,1]==6),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==7,-1],2,mean),
                                                nrow = sum(new_zip[,1]==7),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==8,-1],2,mean),
                                                nrow = sum(new_zip[,1]==8),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T),
                                 matrix(apply(new_zip[new_zip[,1]==9,-1],2,mean),
                                                nrow = sum(new_zip[,1]==9),
                                                ncol = ncol(new_zip[,-1]),
                                                by = T)) %*% reduced_zip$rotation


pairs(reduced_zip$x[,1:3] + z[,1:3],
      panel=function(x,y){panel.smooth(x,y, col = as.numeric(zip2[,1])+2,
        pch = 20, cex = 1) 
        abline(lsfit(x,y),lty=2) })
# print("Reduced PCA Summary:")
# print(summary(reduced_zip))
# plot(reduced_zip$x[,1:2], pch=16, col=as.numeric(as.factor(new_zip[,"digits"])), main="Reduced Handwritten Zipcode PCA separated by Digit") 
# legend(-11, 4, pch=16, col=unique(as.numeric(as.factor(new_zip[,"digits"]))),legend=unique(new_zip[,"digits"]))

```
I wasn't sure how to show the first 32 principal components for the full data set or what the best way to compare the full to reduced dataset was. Here I am showing a comparison of the first 3 principal components between the full and reduced data sets. Both plots were created using digit mean centered data, the only difference is the first used all of the data and the second uses the reduced dataset from part.A)
I think that the reduced shows better separation between the digits, namely 1, 2 and maybe 0 or 8. This is most likely because we reduced the data down to only include the 100  most significant pixels that contribute to the categorization. 


### part Bii.) One aspect of using PCA is to obtain a lower-rank representation of the images. Suppose that we have $\Sigma$ = $\Gamma$ $\Lambda$ $\Gamma$' in terms of the spectral decomposition. Assume that $\Gamma_q$ contains the first q eigenvectors (columns) of $\Sigma$ . Then the variance of $\Gamma_q \Gamma_q' X$ has a lower-rank (q) is than $\Sigma$ and indeed does not contain the smaller eigenvalues (largely random noise, and not a source of variability owing to signal). Using this transformation, display the digits in the raw form (use a 10 × 20 grid) and in terms of the lower-rank form. Comment on the results. [10 points]



```{r}
zip <- as.matrix(zip)
# Raw Data
par(mfrow = c(10,20), mar = rep(0.1, 4))
for (i in 1:2000)
  image(z = matrix(zip[i,], ncol = 16, by = F)[,16:1], axes = F, col = gray(15:0/15))


# Reduced Data
gamma_q <- pca_zip$rotation[,1:32]
gamma_q_t <- t(gamma_q)
test <- gamma_q %*% gamma_q_t %*% t(zip)
test <- t(test)

par(mfrow = c(10,20), mar = rep(0.1, 4))
for (i in 1:2000)
  image(z = matrix(test[i,], ncol = 16, by = F)[,16:1], axes = F, col = gray(16:0/16))

```
I understand that the reduced data is supposed to present a clearer image because we are using principal components and spectral decomposition to reduce the noise of the data. My reduced data does not look clearer to me. THe backgrounds are slightly different, but I used the same gray scale. Although the digits are still readable.  




